\textbf{ Analytic Light Curves in Reflected Light: \\% Phase Curves, Occultations, and Non-Lambertian Scattering for Spherical Planets and Moons }

### 2021-02-26 tex/ms.tex
tweaks

There is an extensive literature on techniques for mapping spherical exoplanet surfaces
based on their phase curves
\citep[e.g.,][]{Russell1906,Lacis1972,Knutson2007,Cowan2008,Oakley2009,Berdyugina2017}
and occultation light curves
\citep[e.g.,][]{Williams2006,Rauscher2007,Majeau2012,deWit2012,Rauscher2018},
both in thermal and reflected (scattered) light.
In particular, much attention has been given to techniques for mapping
Earth-like planets from visible-light reflected phase curves
\citep[e.g.,][]{Ford2001,Kawahara2010,Kawahara2011,Fujii2012,Kawahara2020,Aizawa2020}.
Unlike thermal phase curves, which primarily encode (often degenerate)
information about longitudinal surface brightness variations
\citep{Russell1906}, reflected light curves often contain information
about the full two-dimensional surface albedo distribution
\citep[e.g.,][]{Kawahara2010}.

Occultation light curves in reflected light can encode even more information
about the surface. Thus far, these have been studied primarily within our
solar system. Mutual occultations among the Galilean moons of Jupiter
have been extensively studied to infer surface properties of the moons and to refine
their ephemerides
\citep[e.g.,][]{Arlot1974,Aksnes1984,Arlot2014,deKleer2017,Saquet2018,Morgado2019}.
Farther out in the solar system,
mutual occultations of Pluto and Charon in the late 1980s
were used to confirm Charon's existence
\citep{Stern1992},
establish the sizes and orbital parameters of the two bodies
\citep{Tholen1990},
and infer their surface properties
\citep{Marcialis1990}.
In particular, \citet{Dunbar1986} developed an efficient analytic
algorithm to model Pluto-Charon occultation light curves in reflected light
assuming uniform surfaces and used it to infer the two body's
average geometrical albedos. Later, \citet{Reinsch1994} analyzed
the complete mutual occultation dataset to infer longitudinal maps of Pluto's
albedo.

@@

As in \citet{Luger2019}, we compute fluxes by first expanding the surface
in terms of spherical harmonics.
While in \citet{Luger2019} we expanded the emissivity of the surface, here
we instead expand the spherical albedo $A$.
Specifically, if $\bvec{y}$ is the vector
of spherical harmonic coefficients describing the albedo anywhere on
the surface and
$\by$ is the spherical harmonics basis (Equation~\ref{eq:by}),
the albedo $A$ at a point $(x, y)$ on the sky-projected disk of the body
is given by the dot product
%
\begin{align}
    \label{eq:albedo}
    A(x, y) = \by^\top (x, y) \, \bvec{y}
    \quad.
\end{align}
%
The flux measured from this body is proportional to the surface integral
over the projected disk
of the albedo $A$ times the illumination profile $\mathcal{I}$ of the surface,
given by Lambert's law as
%
\begin{align}
    \label{eq:LambertsLaw}
    \mathcal{I}(\vartheta_i) = \mathcal{I}_0 \, \text{max}\big( 0, \cos\vartheta_i \big)
    \quad,
\end{align}
%
where $\vartheta_i$ is the angle between the incident radiation and the surface
normal and $\mathcal{I}_0$ is the peak illumination. We show in
Appendix~\ref{sec:adapting-starry} that in the case of, say, a planet
illuminated by its host star,
%
\begin{align}
    \mathcal{I}_0 = \frac{f_s}{\pi r_s^2}
\end{align}
%
where $r_s$ is the distance between the planet and the star (in units
of the planet's radius) and $f_s$ is the stellar flux measured at the observer
(in arbitrary units). Following the convention in \citet{Luger2019}, we assume
throughout this paper that $f_s = 1$, so all fluxes are defined as a fraction of
the flux of the illumination source at the observer.

@@

Figure~\ref{fig:earthmoon} shows another light curve of the rotating
Earth, but this time taken during an occultation by the Moon. The map
of the Earth is the same as before, but the observer is now in the
equatorial frame of the Earth.
The top panel is a reproduction of
Figure~7 in \citet{Luger2019} for the case of thermal light,
where the Moon is seen to travel across the
disk of the Earth from southwest to northeast, progressively occulting
South America (dip), the Atlantic (peak), and Africa (dip).
As before,
the blue curve is the analytic solution and the orange dots correspond
to the numerical solution.

@@

In the limit that the angular size of the star as seen from the planet is
small, the illumination profile on the surface of the planet will
decrease as the cosine of the angle between the surface normal and the star,
reaching zero at the day/night terminator, an angle $\nicefrac{\pi}{2}$
away from the sub-stellar point. However, if the star is sufficiently large
and the planet is sufficiently close-in, rays originating from near the limb
of the star will reach points on the planet surface beyond this angle.
If the stellar radius $R_\star$
is larger than the planet radius $R_p$, the angular extent of the true
day/night terminator past $\nicefrac{\pi}{2}$ is given by
%
\begin{proof}{tau}
    \label{eq:tau}
    \tau &= \arcsin\left( \frac{1 - \nicefrac{R_p}{R_\star}}{\nicefrac{a}{R_\star}} \right)
\end{proof}
%
where $a$ the semi-major axis of the orbit (where we implictly assume
the eccentricity is zero).
For planets like the Earth, this quantity is only about $0.26^\circ$,
resulting in a negligible effect on the planet's light curve. However,
for very close-in planets, the effect can be significant. For instance,
the hot Jupiter Kelt-9b has $\nicefrac{R_p}{R_\star} = 0.083$ and
$\nicefrac{a}{R_\star} = 3.16$
planet \citep{Wong2019}. Assuming zero eccentricity and ignoring any
stellar oblateness
\citep[see][]{Ahlers2020},
the day/night terminator extends $\tau \approx 17^\circ$ past the limb
of that planet. Figure~\ref{fig:extended} shows the illumination profile
of this planet in a Mollweide projection, with the sub-stellar point at
the center, for the point source approximation (left) and accounting for
the finite size of the star (right). In addition to the
displaced day/night terminator, the main difference between
the two profiles is the sub-stellar intensity, which is significantly higher
in the extended source case. This is due to the simple fact that in the point
source case the illumination source is placed at the center of the star,
which is one stellar radius farther from the planet than the point closest
to the planet (the sub-planetary point) in the extended source case.
Once accounting for this difference, the fractional change in the intensity
on the planet away from the sub-stellar point is similar in both cases,
and the intensity anywhere beyond $\nicefrac{\pi}{2}$ is less than one-tenth
the peak value.

@@

The illumination profile in the extended source case may be computed as the
two-dimensional integral of the point source illumination profile over
the visible portion of the stellar disk. While this integral may in theory be
computed analytically
\citep[see, for instance,][who derived series solutions to this problem]{Kopal1954},
the resulting profile on the planet surface
will not in general be exactly expressible in terms of spherical harmonics, a
necessary condition for the \starry algorithm.
For simplicity, we therefore compute the illumination profile for
extended sources by averaging the contribution of \textsf{source\_npts} point
sources uniformly distributed across the projected disk corresponding to
the portion of the stellar surface
visible from the planet, where \textsf{source\_npts} is a user-supplied
value. In the limit $R_p \ll R_\star$, this is a spherical cap centered at
the sub-planet point with radius $R_\star \cos\tau$. In
Figure~\ref{fig:extended} we set
\textsf{source\_npts} = 300, but in practice we find that ${\sim}30$
points are sufficient for even the most extreme cases such as Kelt-9b.
%
Note, importantly, that while this method allows one to account for the
effect of stellar limb darkening on the illumination profile of the planet,
this has not been implemented in \starry.

@@

In principle, any of these processes can be accounted for in
the \starry algorithm by modifying the linear operator
$\mathbf{I}$ (Equation~\ref{eq:Imat}), which in the Lambertian
case simply weights the spherical harmonic expansion of the albedo
by the cosine-like illumination profile to obtain the observed
intensity at a point on the surface
(see Equations~\ref{eq:sTA2IA1RRy} and \ref{eq:rTIA1RRy}).
For non-Lambertian scattering, this matrix must also account for
the phase dependence of the reflection: in particular, it will depend
not only on the angle between the surface normal and the incident
radiation, $\vartheta_i$, but also on the angle between the surface
normal and the reflected radiation (i.e., the direction toward the
observer), $\vartheta_r$. It may also depend on the azimuthal angles
of the incident and reflected rays, $\phi_i$ and $\phi_r$, respectively.
These four angles are shown in Figure~\ref{fig:scattering}, showing
the incoming radiation (source) vector $\mathbf{s}$ and the outgoing
radiation (viewer) vector $\mathbf{v}$ in a frame in which the $z$-axis
points along the surface normal.

Treatment of a generalized, flexible scattering model is beyond the scope
of this paper. However, as an example of how a scattering model may be
incorporated
into the \starry algorithm, we consider in detail the case of the rough surface
scattering model of \citet{OrenNayar1994}, commonly used in computer
graphics applications and solar system body modeling
\citep[e.g.,][]{Morgado2019}. In this model, the surface is treated as a
collection of a large number of Lambertian facets oriented at random
angles relative to the average surface normal, whose net contribution to
the total intensity can depart significantly from the Lambertian case.
While the general model accounts for interreflections, shadowing, and
an arbitrary distribution of facet orientations, in its simplest form
the intensity observed at a point $(x, y)$ on the (projected) surface of a
body of unit spherical albedo may be approximated as
\citep[c.f. Equation~30 in][]{OrenNayar1994}
%
\begin{proof}{OrenNayar}
    \label{eq:OrenNayar}
    \mathcal{I} & =
    \mathcal{I}_\text{Lamb}
    \bigg\{ c_0 + c_1 \, \text{max}\Big( 0, \, \cos(\phi_r - \phi_i) \Big)
    \sin \alpha \tan \beta
    \bigg\}
\end{proof}
%
where%
%
\\
\begin{minipage}{.5\linewidth}
    \begin{align}
        c_0 & = 1 - 0.5 \left(\frac{\sigma^2}{\sigma^2 + 0.33}\right)
        \nonumber                                                     \\
        c_1 & = 0.45 \left(\frac{\sigma^2}{\sigma^2 + 0.09}\right)
        \nonumber
    \end{align}
\end{minipage}%
\begin{minipage}{.49\linewidth}
    \begin{proof}{OrenNayar}
        \alpha & = \text{max}\Big( \vartheta_r, \, \vartheta_i \Big)
        \nonumber                                                    \\
        \beta  & = \text{min}\Big( \vartheta_r, \, \vartheta_i \Big)
    \end{proof}
\end{minipage}
\\[1em]
and the angles $\vartheta_i$, $\vartheta_r$, $\phi_i$, and $\phi_r$ are all
implicit functions of $x$, $y$, and the illumination source position.
The term $\mathcal{I}_\text{Lamb}$ is the Lambertian illumination profile,
given by Equation~(\ref{eq:LambertsLaw}).
%
At a given point on the surface, and for a given source position,
the intensity $\mathcal{I}$ is therefore a function of a single
parameter, $\sigma$, defined as the standard deviation in radians of the
distribution of facet angles (which is assumed to be a zero-mean Gaussian).

@@

In order to incorporate this scattering model into \starry, we must weight
the spherical harmonic expansion of the albedo, $\mathbf{y}$, by
Equation~(\ref{eq:OrenNayar}) instead of Equation~(\ref{eq:LambertsLaw}).
Weighting by Equation~(\ref{eq:LambertsLaw}) is (relatively) straightforward,
since $\mathcal{I}_\text{Lamb}$ is a piecewise function of the $l=1$ spherical
harmonics (see Appendix~\ref{sec:adapting-starry}).
The function we must integrate when computing fluxes is therefore
exactly expressible in terms of spherical harmonics and thus
\starry-integrable. However, Equation~(\ref{eq:OrenNayar}) cannot be expressed
exactly in terms of spherical harmonics, so we must instead approximate it.
To this end, we evaluate Equation~(\ref{eq:OrenNayar}) on a
grid of $x$ and $y$ spanning the unit disk, as well as the illumination phase,
parametrized by $b$, the semi-minor axis of the elliptical segment defining
the day/night terminator (see Appendix~\ref{sec:adapting-starry}).
We then fit to this a polynomial of total degree \STARRYORENNAYARDEG
in $x$, $y$, and $z \equiv \sqrt{1 - x^2 - y^2}$ and degree
\STARRYORENNAYARNB in $b$ and degree \STARRYORENNAYARNBC in
$b_c \equiv \sqrt{1 - b^2}$.
%
Then, for a given value of $b$ and $b_c$, we construct the operator
$\mathbf{I}$ out of the polynomial coefficients in the same way as we
constructed the Lambertian operator in Appendix~\ref{sec:adapting-starry}.
%
More details about our approximation can be found in the Jupyter notebook
accompanying Equation~(\ref{eq:OrenNayar}).

@@

Our implementation of the scattering model extends just as easily
to occultations and to cases where the surface does not have uniform albedo.
Figure~\ref{fig:io_europa} shows a visible-light observation of the
occultation of Io by Europa on 04 Dec 2009 taken by the PHEMU09
campaign \citep{Arlot2014}. The trajectory of Europa relative to Io,
computed from the \href{https://ssd.jpl.nasa.gov/horizons.cgi}{JPL Horizons database}
using ephemerides from \citet{Folkner2014}
is shown at the top. We fit to this data a \starry occultation model
with the scattering law discussed above.
For simplicity, we set the surface map equal too
an $l=15$ spherical harmonic expansion of the Galileo global color mosaic
of Io \citep{Becker2005} and use ephemerides from the JPL Horizons
database, allowing for a small static $x-y$ offset between Europa and Io as
in \citep{Arlot2014} due to the uncertainty in the database. In total,
we fit for five parameters: the two Cartesian offset terms,
the flux contribution from Europa, the average
albedo of Io, and the average surface roughness of Io, $\sigma$. The model
is displayed in blue and closely matches the data.
%
Note, however, that this is meant simply as a demonstration of the \starry
algorithm, as our model for the surface is approximate at best, given
differences in the wavelength band between the Galileo observations and
those of the PHEMU09 campaign, changes in the albedo of Io since the Galileo
measurements, and the fact that the orientation and extent of any shadows
due to volcanoes on the surface are likely different between the Galileo
and PHEMU09 observations. Furthermore,
proper modeling would entail the
joint analysis of all light curves of Io taken in a given season, for which
we can afford to simultaneously fit for the surface map without risk of
overfitting (Bartoli\'c, Luger et al., \emph{in prep}).

@@

One of the fundamental difficulties with the mapping problem is that
the process of inferring a two-dimensional map from a light curve
is almost always ill-posed. This has been known since at least
the turn of the last century, when \citet{Russell1906} discussed
how, because of symmetry,
all odd harmonics above $l = 1$ are in the null space for thermal
phase curves of spherical bodies, meaning those terms do not contribute
at all to the disk-integrated flux. As discussed in \citet{Luger2019},
the problem is even more ill-posed than that: for any even
degree $l > 0$, there are $2l + 1$ modes on the \emph{surface}
(one for each value of $m$), but only $2$ Fourier modes in the
\emph{light curve} (i.e., a sine and a cosine). Thus, for every
mode that can be constrained from the light curve, there are far
more modes that cannot, a problem that only gets worse as $l$
increases.

@@

There are several limitations to the \starry model that are important
to bear in mind. The primary limitation concerns the maximum spherical
harmonic degree of the model. While the expressions derived here are
valid at arbitrary degree, we find that their numerical stability begins
to degrade above $l \sim 20-25$ for occultations and $l \sim 35-40$ for phase
curves. The same is true for the model in thermal light \citep{Luger2019},
and is due to (1) the large condition number of the change-of-basis matrix
from spherical harmonics to polynomials and (2) instabilities in the
many recursion relations used to evaluate the solution vectors
$\sT$ and $\rT$.
In principle, one could improve the numerical stability by evaluating
all expressions at higher floating-point precision, but in practice
the computational cost of this becomes quickly prohibitive.
However, it is important to keep in mind that our current best image
of an exoplanet is the $l=1$ map of HD189733b
\citep{Knutson2007,Majeau2012,deWit2012}. Next-generation facilities
such as JWST may allow us to probe surface modes as small as $l=5$
for some planets (Luger et al., \emph{in prep}), but even with future
telescopes such as LUVOIR it is extremely unlikely we will do better than
$l=20$. If cases arise requiring a resolution smaller than about
$180^\circ / 20 = 9^\circ$ on the surface, the \starry algorithm
will have to be revisited.

@@

The third limitation concerns the flexibility of the \starry model.
While we presented ways to capture non-Lambertian scattering
in \starry (\S\ref{sec:nonlambertian}), there are certain aspects
of light curves in reflected light that cannot be captured by the model.
One example of this is shadowing. Craters on the moon or
volcanoes on Io can cast large shadows visible from space, particularly
if viewed near crescent phase. Unfortunately, there is no way to model
this within the \starry framework. Another example is multiple scattering,
as in optically thick atmospheres, for which a proper radiative transfer model
must be used. There is also the case of specular reflection,
or ``glint'', which is a pronounced signal for the Earth due its
oceans \citep{Robinson2014} and on Titan due to
its hydrocarbon lakes \citep{Barnes2011}. In principle, glint could be
modeled in the same way as non-Lambertian sccattering, by constructing the
linear operator $\mathbf{I}$ in such a way as to downweight portions of the
projected disk where $\vartheta_i \neq \vartheta_r$. In practice, however,
if the size of the glint spot is small (which is typically the case),
an expansion at extremely high $l$ ($l \sim 360$ in the case of the Earth)
would be required, which would not work for the reasons above.
Instead, it may be possible to combine the \starry algorithm with the
formalism of \citet{Haggard2018}, who derived analytic expressions for
phase curves of delta function maps ($\delta$-map), to model glint.
As discussed in \citet{LustigYaeger2018}, glint mapping is an extremely
powerful way to not only map terrestrial planets but also to confirm
their habitability via the presence of an ocean.  Lastly, the presence
of time-variable features, such as clouds, dust storms, or seasonal
variations of vegetation are not accounted for in the model, although
the \href{https://starry.readthedocs.io}{documentation} discusses
how one may approach the modeling of temporal features.

>>>>

There is an extensive literature on techniques for mapping exoplanet surfaces
based on their phase curves
\citep[e.g.,][]{Russell1906,Lacis1972,Knutson2007,Cowan2008,Oakley2009,Berdyugina2017,PaperI,PaperII}
and occultation light curves
\citep[e.g.,][]{Williams2006,Rauscher2007,Majeau2012,deWit2012,Rauscher2018},
both in thermal and reflected (scattered) light.
In particular, much attention has been given to techniques for mapping
Earth-like planets from visible-light reflected phase curves
\citep[e.g.,][]{Ford2001,Kawahara2010,Kawahara2011,Fujii2012,Kawahara2020,Aizawa2020}.
Unlike thermal phase curves, which primarily encode (often degenerate)
information about longitudinal surface brightness variations
\citep{Russell1906}, reflected light curves often contain information
about the full two-dimensional surface albedo distribution
\citep[e.g.,][]{Kawahara2010}.

Occultation light curves in reflected light can encode even more information
about the surface. Thus far, these have been studied primarily within our
solar system. Mutual occultations among the Galilean moons of Jupiter
have been extensively studied to infer surface properties of the moons and to refine
their ephemerides
\citep[e.g.,][]{Arlot1974,Aksnes1984,Arlot2014,deKleer2017,Saquet2018,Morgado2019,Bartolic2021}.
Farther out in the solar system,
mutual occultations of Pluto and Charon in the late 1980s
were used to confirm Charon's existence
\citep{Stern1992},
establish the sizes and orbital parameters of the two bodies
\citep{Tholen1990},
and infer their surface properties
\citep{Marcialis1990}.
In particular, \citet{Dunbar1986} developed an efficient analytic
algorithm to model Pluto-Charon occultation light curves in reflected light
assuming uniform surfaces and used it to infer the two body's
average geometrical albedos. Later, \citet{Reinsch1994} analyzed
the complete mutual occultation dataset to infer longitudinal maps of Pluto's
albedo.

@@

As in \citet{Luger2019}, we compute fluxes by first expanding the surface
in terms of spherical harmonics.
While in \citet{Luger2019} we expanded the emissivity of the surface, here
we instead expand the spherical albedo $A$.
Specifically, if $\bvec{y}$ is the vector
of spherical harmonic coefficients describing the albedo anywhere on
the surface and
$\by$ is the spherical harmonics basis (Equation~\ref{eq:by}),
the albedo $A$ at a point $(x, y)$ on the sky-projected disk of the body
is given by the dot product
%
\begin{align}
    \label{eq:albedo}
    A(x, y) = \by^\top (x, y) \, \bvec{y}
    \quad.
\end{align}
%
The flux measured from this body is proportional to the surface integral
over the projected disk
of the albedo $A$ times the illumination profile $\mathcal{I}$ of the surface,
given by Lambert's law as
%
\begin{align}
    \label{eq:LambertsLaw}
    \mathcal{I}(\vartheta_\mathrm{i}) = \mathcal{I}_0 \, \text{max}\big( 0, \cos\vartheta_\mathrm{i} \big)
    \quad,
\end{align}
%
where $\vartheta_\mathrm{i}$ is the angle between the incident radiation and the surface
normal and $\mathcal{I}_0$ is the peak illumination. We show in
Appendix~\ref{sec:adapting-starry} that in the case of, say, a planet
illuminated by its host star,
%
\begin{align}
    \mathcal{I}_0 = \frac{f_s}{\pi r_\mathrm{s}^2}
\end{align}
%
where $r_\mathrm{s}$ is the distance between the planet and the star (in units
of the planet's radius) and $f_s$ is the stellar flux measured at the observer
(in arbitrary units). Following the convention in \citet{Luger2019}, we assume
throughout this paper that $f_s = 1$, so all fluxes are defined as a fraction of
the flux of the illumination source at the observer.

@@

Figure~\ref{fig:earthmoon} shows another light curve of the rotating
Earth, but this time taken during an occultation by the Moon. The map
of the Earth is the same as before, but the observer is now along the
equatorial plane of the Earth.
The top panel is a reproduction of
Figure~7 in \citet{Luger2019} for the case of thermal light,
where the Moon is seen to travel across the
disk of the Earth from southwest to northeast, progressively occulting
South America (dip), the Atlantic (peak), and Africa (dip).
As before,
the blue curve is the analytic solution and the orange dots correspond
to the numerical solution.

@@

In the limit that the angular size of the star as seen from the planet is
small, the illumination profile on the surface of the planet will
decrease as the cosine of the angle between the surface normal and the star,
reaching zero at the day/night terminator, an angle $\nicefrac{\pi}{2}$
away from the sub-stellar point. However, if the star is sufficiently large
and the planet is sufficiently close-in, rays originating from near the limb
of the star will reach points on the planet surface beyond this angle.
If the stellar radius $R_\star$
is larger than the planet radius $R_\mathrm{p}$, the angular extent of the true
day/night terminator past $\nicefrac{\pi}{2}$ is given by
%
\begin{proof}{tau}
    \label{eq:tau}
    \tau &= \arcsin\left( \frac{1 - \nicefrac{R_\mathrm{p}}{R_\star}}{\nicefrac{a}{R_\star}} \right)
\end{proof}
%
where $a$ the semi-major axis of the orbit (where we implictly assume
the eccentricity is zero).
For planets like the Earth, this quantity is only about $0.26^\circ$,
resulting in a negligible effect on the planet's light curve. However,
for very close-in planets, the effect can be significant. For instance,
the hot Jupiter Kelt-9b has $\nicefrac{R_\mathrm{p}}{R_\star} = 0.083$ and
$\nicefrac{a}{R_\star} = 3.16$
planet \citep{Wong2019}. Assuming zero eccentricity and ignoring any
stellar oblateness
\citep[see][]{Ahlers2020},
the day/night terminator extends $\tau \approx 17^\circ$ past the limb
of that planet. Figure~\ref{fig:extended} shows the illumination profile
of this planet in a Mollweide projection, with the sub-stellar point at
the center, for the point source approximation (left) and accounting for
the finite size of the star (right). In addition to the
displaced day/night terminator, the main difference between
the two profiles is the sub-stellar intensity, which is significantly higher
in the extended source case. This is due to the simple fact that in the point
source case the illumination source is placed at the center of the star,
which is one stellar radius farther from the planet than the point closest
to the planet (the sub-planetary point) in the extended source case.
Once accounting for this difference, the fractional change in the intensity
on the planet away from the sub-stellar point is similar in both cases,
and the intensity anywhere beyond $\nicefrac{\pi}{2}$ is less than one-tenth
the peak value.

@@

The illumination profile in the extended source case may be computed as the
two-dimensional integral of the point source illumination profile over
the visible portion of the stellar disk. While this integral may in theory be
computed analytically
\citep[see, for instance,][who derived series solutions to this problem]{Kopal1954},
the resulting profile on the planet surface
will not in general be exactly expressible in terms of spherical harmonics, a
necessary condition for the \starry algorithm.
For simplicity, we therefore compute the illumination profile for
extended sources by averaging the contribution of \textsf{source\_npts} point
sources uniformly distributed across the projected disk corresponding to
the portion of the stellar surface
visible from the planet, where \textsf{source\_npts} is a user-supplied
value. In the limit $R_\mathrm{p} \ll R_\star$, this is a spherical cap centered at
the sub-planet point with radius $R_\star \cos\tau$. In
Figure~\ref{fig:extended} we set
\textsf{source\_npts} = 300, but in practice we find that ${\sim}30$
points are sufficient for even the most extreme cases such as Kelt-9b.
%
Note, importantly, that while this method allows one to account for the
effect of stellar limb darkening on the illumination profile of the planet,
this has not been implemented in \starry.

@@

In principle, any of these processes can be accounted for in
the \starry algorithm by modifying the linear operator
$\mathbf{I}$ (Equation~\ref{eq:Imat}), which in the Lambertian
case simply weights the spherical harmonic expansion of the albedo
by the cosine-like illumination profile to obtain the observed
intensity at a point on the surface
(see Equations~\ref{eq:sTA2IA1RRy} and \ref{eq:rTIA1RRy}).
For non-Lambertian scattering, this matrix must also account for
the phase dependence of the reflection: in particular, it will depend
not only on the angle between the surface normal and the incident
radiation, $\vartheta_\mathrm{i}$, but also on the angle between the surface
normal and the reflected radiation (i.e., the direction toward the
observer), $\vartheta_\mathrm{r}$. It may also depend on the azimuthal angles
of the incident and reflected rays, $\phi_\mathrm{i}$ and $\phi_\mathrm{r}$, respectively.
These four angles are shown in Figure~\ref{fig:scattering}, showing
the incoming radiation (source) vector $\mathbf{s}$ and the outgoing
radiation (viewer) vector $\mathbf{v}$ in a frame in which the $z$-axis
points along the surface normal.

Treatment of a generalized, flexible scattering model is beyond the scope
of this paper. However, as an example of how a scattering model may be
incorporated
into the \starry algorithm, we consider in detail the case of the rough surface
scattering model of \citet{OrenNayar1994}, commonly used in computer
graphics applications and solar system body modeling
\citep[e.g.,][]{Morgado2019}. In this model, the surface is treated as a
collection of a large number of Lambertian facets oriented at random
angles relative to the average surface normal, whose net contribution to
the total intensity can depart significantly from the Lambertian case.
While the general model accounts for interreflections, shadowing, and
an arbitrary distribution of facet orientations, in its simplest form
the intensity observed at a point $(x, y)$ on the (projected) surface of a
body of unit spherical albedo may be approximated as
\citep[c.f. Equation~30 in][]{OrenNayar1994}
%
\begin{proof}{OrenNayar}
    \label{eq:OrenNayar}
    \mathcal{I} & =
    \mathcal{I}_\text{Lamb}
    \bigg\{ c_0 + c_1 \, \text{max}\Big( 0, \, \cos(\phi_\mathrm{r} - \phi_\mathrm{i}) \Big)
    \sin \alpha \tan \beta
    \bigg\}
\end{proof}
%
where%
%
\\
\begin{minipage}{.5\linewidth}
    \begin{align}
        c_0 & = 1 - 0.5 \left(\frac{\sigma^2}{\sigma^2 + 0.33}\right)
        \nonumber                                                     \\
        c_1 & = 0.45 \left(\frac{\sigma^2}{\sigma^2 + 0.09}\right)
        \nonumber
    \end{align}
\end{minipage}%
\begin{minipage}{.49\linewidth}
    \begin{proof}{OrenNayar}
        \alpha & = \text{max}\Big( \vartheta_\mathrm{r}, \, \vartheta_\mathrm{i} \Big)
        \nonumber                                                    \\
        \beta  & = \text{min}\Big( \vartheta_\mathrm{r}, \, \vartheta_\mathrm{i} \Big)
    \end{proof}
\end{minipage}
\\[1em]
and the angles $\vartheta_\mathrm{i}$, $\vartheta_\mathrm{r}$, $\phi_\mathrm{i}$, and $\phi_\mathrm{r}$ are all
implicit functions of $x$, $y$, and the illumination source position.
The term $\mathcal{I}_\text{Lamb}$ is the Lambertian illumination profile,
given by Equation~(\ref{eq:LambertsLaw}).
%
At a given point on the surface, and for a given source position,
the intensity $\mathcal{I}$ is therefore a function of a single
parameter, $\sigma$, defined as the standard deviation in radians of the
distribution of facet angles (which is assumed to be a zero-mean Gaussian).

@@

In order to incorporate this scattering model into \starry, we must weight
the spherical harmonic expansion of the albedo, $\mathbf{y}$, by
Equation~(\ref{eq:OrenNayar}) instead of Equation~(\ref{eq:LambertsLaw}).
Weighting by Equation~(\ref{eq:LambertsLaw}) is (relatively) straightforward,
since $\mathcal{I}_\text{Lamb}$ is a piecewise function of the $l=1$ spherical
harmonics (see Appendix~\ref{sec:adapting-starry}).
The function we must integrate when computing fluxes is therefore
exactly expressible in terms of spherical harmonics and thus
\starry-integrable. However, Equation~(\ref{eq:OrenNayar}) cannot be expressed
exactly in terms of spherical harmonics, so we must instead approximate it.
To this end, we evaluate Equation~(\ref{eq:OrenNayar}) on a
grid of $x$ and $y$ spanning the unit disk, as well as the illumination phase,
parametrized by $b$, the semi-minor axis of the elliptical segment defining
the day/night terminator (see Appendix~\ref{sec:adapting-starry}).
We then fit to this a polynomial of total degree \STARRYORENNAYARDEG
in $x$, $y$, and $z \equiv \sqrt{1 - x^2 - y^2}$ and degree
\STARRYORENNAYARNB in $b$ and degree \STARRYORENNAYARNBC in
$b_\mathrm{c} \equiv \sqrt{1 - b^2}$.
%
Then, for a given value of $b$ and $b_\mathrm{c}$, we construct the operator
$\mathbf{I}$ out of the polynomial coefficients in the same way as we
constructed the Lambertian operator in Appendix~\ref{sec:adapting-starry}.
%
More details about our approximation can be found in the Jupyter notebook
accompanying Equation~(\ref{eq:OrenNayar}).

@@

Our implementation of the scattering model extends just as easily
to occultations and to cases where the surface does not have uniform albedo.
Figure~\ref{fig:io_europa} shows a visible-light observation of the
occultation of Io by Europa on 04 Dec 2009 taken by the PHEMU09
campaign \citep{Arlot2014}. The trajectory of Europa relative to Io,
computed from the \href{https://ssd.jpl.nasa.gov/horizons.cgi}{JPL Horizons database}
using ephemerides from \citet{Folkner2014}
is shown at the top. We fit to this data a \starry occultation model
with the scattering law discussed above.
For simplicity, we set the surface map equal too
an $l=15$ spherical harmonic expansion of the Galileo global color mosaic
of Io \citep{Becker2005} and use ephemerides from the JPL Horizons
database, allowing for a small static $x-y$ offset between Europa and Io as
in \citep{Arlot2014} due to the uncertainty in the database. In total,
we fit for five parameters: the two Cartesian offset terms,
the flux contribution from Europa, the average
albedo of Io, and the average surface roughness of Io, $\sigma$. The model
is displayed in blue and closely matches the data.
%
Note, however, that this is meant simply as a demonstration of the \starry
algorithm, as our model for the surface is approximate at best, given
differences in the wavelength band between the Galileo observations and
those of the PHEMU09 campaign, changes in the albedo of Io since the Galileo
measurements, and the fact that the orientation and extent of any shadows
due to volcanoes on the surface are likely different between the Galileo
and PHEMU09 observations. Furthermore,
proper modeling would entail the
joint analysis of all light curves of Io taken in a given season, for which
we can afford to simultaneously fit for the surface map without risk of
overfitting \citep{Bartolic2021}.

@@

One of the fundamental difficulties with the mapping problem is that
the process of inferring a two-dimensional map from a light curve
is almost always ill-posed. This has been known since at least
the turn of the last century, when \citet{Russell1906} discussed
how, because of symmetry,
all odd harmonics above $l = 1$ are in the null space for thermal
phase curves of spherical bodies, meaning those terms do not contribute
at all to the disk-integrated flux. As discussed in \citet{Luger2019}
and \citet{PaperI},
the problem is even more ill-posed than that: for any even
degree $l > 0$, there are $2l + 1$ modes on the \emph{surface}
(one for each value of $m$), but only $2$ Fourier modes in the
\emph{light curve} (i.e., a sine and a cosine). Thus, for every
mode that can be constrained from the light curve, there are far
more modes that cannot, a problem that only gets worse as $l$
increases.

@@

There are several limitations to the \starry model that are important
to bear in mind. The primary limitation concerns the maximum spherical
harmonic degree of the model. While the expressions derived here are
valid at arbitrary degree, we find that their numerical stability quickly
degrades above $l \sim 20-25$ for occultations and $l \sim 35-40$ for phase
curves. The same is true for the model in thermal light \citep{Luger2019},
and is due to (1) the large condition number of the change-of-basis matrix
from spherical harmonics to polynomials and (2) instabilities in the
many recursion relations used to evaluate the solution vectors
$\sT$ and $\rT$.
In principle, one could improve the numerical stability by evaluating
all expressions at higher floating-point precision, but in practice
the computational cost of this becomes quickly prohibitive.
However, it is important to keep in mind that our current best image
of an exoplanet is the $l=1$ map of HD189733b
\citep{Knutson2007,Majeau2012,deWit2012}. Next-generation facilities
such as JWST may allow us to probe surface modes as small as $l=5$
for some planets (Luger et al., \emph{in prep}), but even with future
telescopes such as LUVOIR it is extremely unlikely we will do better than
$l=20$. If cases arise requiring a resolution smaller than about
$180^\circ / 20 = 9^\circ$ on the surface, the \starry algorithm
will have to be revisited.

@@

The third limitation concerns the flexibility of the \starry model.
While we presented ways to capture non-Lambertian scattering
in \starry (\S\ref{sec:nonlambertian}), there are certain aspects
of light curves in reflected light that cannot be captured by the model.
One example of this is shadowing. Craters on the moon or
volcanoes on Io can cast large shadows visible from space, particularly
if viewed near crescent phase. Unfortunately, there is no way to model
this within the \starry framework. Another example is multiple scattering,
as in optically thick atmospheres, for which a proper radiative transfer model
must be used. There is also the case of specular reflection,
or ``glint'', which is a pronounced signal for the Earth due its
oceans \citep{Robinson2014} and on Titan due to
its hydrocarbon lakes \citep{Barnes2011}. In principle, glint could be
modeled in the same way as non-Lambertian sccattering, by constructing the
linear operator $\mathbf{I}$ in such a way as to downweight portions of the
projected disk where $\vartheta_\mathrm{i} \neq \vartheta_\mathrm{r}$. In practice, however,
if the size of the glint spot is small (which is typically the case),
an expansion at extremely high $l$ ($l \sim 360$ in the case of the Earth)
would be required, which would not work for the reasons above.
Instead, it may be possible to combine the \starry algorithm with the
formalism of \citet{Haggard2018}, who derived analytic expressions for
phase curves of delta function maps ($\delta$-map), to model glint.
As discussed in \citet{LustigYaeger2018}, glint mapping is an extremely
powerful way to not only map terrestrial planets but also to confirm
their habitability via the presence of an ocean.  Lastly, the presence
of time-variable features, such as clouds, dust storms, or seasonal
variations of vegetation are not accounted for in the model, although
the \href{https://starry.readthedocs.io}{documentation} discusses
how one may approach the modeling of temporal features.
